\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-27 paper/algorithm.tex

Our optimizer searches (\Cref{sec:planning}) a restricted space (\Cref{sec:search-space}) of query plans containing imputation operators (\Cref{sec:operators}) for an optimal plan given an estimate of the runtime of the query, information lost through imputation operations, and a parameter trading off the two.

This plan is subject to two constraints (\Cref{sec:placement}).
First, it must not emit any tuples which contain null values, regardless of the state of the base tables.
Second, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) must never observe a null value in any attribute that they operate on directly.
Our planning algorithm is agnostic to the type of imputation used (\Cref{sec:imputation}), as the properties of the imputation are summarized by the cost model (\Cref{sec:cost-model}).

\subsection{Imputation operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all null values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
\textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no null values for
attributes in $C$.

@@

\item A grouping \& aggregation operator is optional, and if present will be performed after the final join.

@@

In order to pick a final plan from the frontier, our model introduces a parameter $\alpha$ which encodes an upper bound on the $\textsc{Loss}$ value that the user is willing
to tolerate relative to the minimum possible amongst frontier plans.

@@

Scan costs are estimated based on the number of tuples, 
page size, and an unit-less IO cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations. We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the IO cost
per page.

@@

In general, evaluating the time complexity for different \textit{Impute} operators is
complicated, as the properties of
different algorithms can vary significantly, the underlying IO cost
is not easily extracted (especially in the case that the entire stream of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). Indeed, the \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans).

@@

\subsection{Complexity}
Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our planning algorithm still operates in exponential time. 
If we removed our restriction on imputation operators to consider not just minimal versus maximal imputations, and instead allowed any arbitrary subset
of attributes to be imputed at a plan node (including those not operated on by that node), then every single operator in the query plan would have a
number of plans exponential in the number of dirty columns.

>>>>

\todobox{this jumps around too much}{
Our optimizer searches (\Cref{sec:planning}) a restricted space (\Cref{sec:search-space}) of query plans containing imputation operators (\Cref{sec:operators}) for an optimal plan given an estimate of the runtime of the query, information lost through imputation operations, and a parameter trading off the two.

This plan is subject to two constraints (\Cref{sec:placement}).
First, it must not emit any tuples which contain null values, regardless of the state of the base tables.
Second, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) must never observe a null value in any attribute that they operate on directly.
Our planning algorithm is agnostic to the type of imputation used (\Cref{sec:imputation}), as the properties of the imputation are summarized by the cost model (\Cref{sec:cost-model}).}

\subsection{Imputation operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all null values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
In our experiments, we instantiate \textit{Impute} with the CE-CART\cite{burgette2010multiple} algorithm, but we highlight that
our system is agnostic to this choice.
\textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no null values for
attributes in $C$.

@@

\item A grouping and aggregation operator is optional and if present will be performed after the final join.

@@

In order to pick a final plan from the frontier, our model introduces a tradeoff parameter $\alpha$ which encodes an upper bound on the $\textsc{Loss}$ value that the user is willing
to tolerate relative to the minimum possible amongst frontier plans.

@@

Scan costs are estimated based on the number of tuples, 
page size, and an unit-less I/O cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations. We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the I/O cost
per page.

@@

In general, evaluating the time complexity for different \textit{Impute} operators is
complicated, as the properties of
different algorithms can vary significantly, the underlying I/O cost
is not easily extracted (especially in the case that the entire stream of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). Indeed, the \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans).

@@

\subsection{Complexity}
\todo{Stengthen complexity argument, as per Sam's feedback}
Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our planning algorithm still operates in exponential time. 
If we removed our restriction on imputation operators to consider not just minimal versus maximal imputations, and instead allowed any arbitrary subset
of attributes to be imputed at a plan node (including those not operated on by that node), then every single operator in the query plan would have a
number of plans exponential in the number of dirty columns.

### 2017-02-27 paper/experiments.tex

\section{Experiments}\label{sec:experiments}
To evaluate the performance of our approach,  we implemented
a prototype system in Java, following a traditional iterator model.
For our experiments we plan and execute queries
for two separate survey-based data sets, showing that our system
is ideal for early dataset exploration.

\subsection{Data sets} \label{subsec:datasets}
\subsubsection{Center for Disease Control: Surveys}
For our first set of experiments, we use survey data collected by the 
Center for Disease Control and Prevention (CDC) in the United States. In particular, we
experiment on a set of tables collected as part of the National
Health and Nutrition Examination Survey (NHANES), a series of studies
conducted by the CDC on a national sample of several thousand individuals\cite{cdc-data}.
The data consists on survey responses, physical examinations, and laboratory
results, amongst others. Our dataset corresponds to the 2013-2014 NHANES
and is available for direct download on the CDC's organization site on 
kaggle.com, a popular data science website.

There are a total of 6 tables associated with the NHANES data set. For purposes
of our experiments, we use three tables:

@@

\begin{table}
    \centering
    \begin{subtable}{0.5\textwidth}
        \centering
       \input{tables/cdc_demo}
        \caption{Demographics}
    \end{subtable}
    \begin{subtable}{0.5\textwidth}
        \centering
        \input{tables/cdc_labs}
        \caption{Laboratory Results}
    \end{subtable}
       \begin{subtable}{0.5\textwidth}
        \centering
        \input{tables/cdc_exams}
        \caption{Physical Results}
    \end{subtable}
    \caption{Missing values in CDC NHANES 2013-2014 data}
    \label{table:nhanes-description} 
\end{table}

@@

For our purposes, we use a version of the data that has been 
pre-processed, but where plenty of missing values remain. It is
worth noting that given the table design (where many attributes
are related), many of these missing values are to be
expected. The dataset we started with had 15,620 rows and 113
columns. We chose 17 attributes, which are outlined in~\Cref{table:fcc-description}, along with the percentage of missing values.

\begin{table}
   \centering
    \input{tables/fcc}
    \caption{Missing values in FCC Survey Data}
   \label{table:fcc-description} 
\end{table}

@@

\subsection{Queries}
We collect a set of queries (\Cref{fig:queries}) that we think are both representative,
in that they could reasonably be written by a user in the course of data analysis, and
interesting to plan.

@@

Queries 5-8 are on the FCC data. In~\Cref{q5}, we calculate the average income for
survey participants, based on their bootcamp attendance. In~\Cref{q6} estimates the
average age of women from the United States who participated. In~\Cref{q7}, we
calculate the average amount of money survey participants with student debt destined to
learning based on their school degree. In~\Cref{q8}, we join the FCC data with a
reference table provided by the World Bank which summarizes GDP per-capita across various
countries\cite{worldbank-data}. The query calculates the average GDP per-capita for
participants based on whether or not they attended a bootcamp.

@@

\begin{table*}
\centering
 \begin{subtable}{\linewidth}
  \input{queries-cdc}
  \caption{Queries on CDC data}
  \label{fig:queries-cdc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{queries-fcc}
 \caption{Queries on FCC data}
 \label{fig:queries-fcc}
 \end{subtable}
  \caption{Queries used in our experiments.}
  \label{fig:queries}
\end{table*}

@@

\Cref{fig:runtimes} provides a summary of the performance results. The first plot
shows average running times for \ProjectName{} plans in two configurations:
quality-optimized and runtime-optimized. The second plot shows, for comparison, the average
running times required for each query when performing imputation at the base tables.  In all
cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
with performance improvement on a factor of ten. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

\begin{figure}
\begin{subfigure}{\linewidth}
\includegraphics[width=\columnwidth]{figures/running_times_imputedb.png}
\caption{Query runtimes with \ProjectName{}}
\label{subfig:project-runtime-queries}
\end{subfigure}
~
\begin{subfigure}{\linewidth}
\includegraphics[width=\columnwidth]{figures/running_times_base_tables.png}
\caption{Query runtimes with imputation on base tables}
\label{subfig:base-runtime-queries}
\end{subfigure}
\caption{Comparing \ProjectName{} runtimes for each query
    (\Cref{subfig:project-runtime-queries}) with the runtime for a query when base-table
    imputations are used (\Cref{subfig:base-runtime-queries})}
\label{fig:runtimes}
\end{figure}

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
\ProjectName{}'s query results when compared to running imputation on the base tables and
executing the query on the cleaned data. Each query was run 200 iterations in both settings
and then results from the two approaches were paired up and compared tuple-wise. We average
tuple-wise absolute percentage deviations within each iteration of a query, and we report
this value averaged over all iterations.  We can see that optimizing for quality indeed
reduces the SMAPE of query results. In general, the SMAPE relative to the base-imputation
approach are low in all cases, indicating that on-the-fly imputation produces similar
results to imputation at the base tables.

>>>>

\section{Experiments}\label{sec:experiments}
To evaluate the performance of our approach,  we implemented
a prototype system in Java, following a traditional iterator model.
For our experiments we plan and execute queries
for two separate survey-based data sets, showing that our system
is well suited for early dataset exploration.

\subsection{Data sets} \label{subsec:datasets}
\subsubsection{Center for Disease Control: Surveys}
We use survey data collected by the 
Center for Disease Control and Prevention (CDC) in the United States. We
experiment on a set of tables collected as part of the 2013-2014 National
Health and Nutrition Examination Survey (NHANES), a series of studies
conducted by the CDC on a national sample of several thousand individuals\cite{cdc-data}.
The data consists on survey responses, physical examinations, and laboratory
results, amongst others. Our dataset is available for direct download on the CDC's organization site on 
kaggle.com, a popular data science website.

There are a total of 6 tables associated with the NHANES data set. We use three tables:

@@

\begin{table}
    \centering
    \begin{subtable}{0.5\textwidth}
        \centering
       \input{tables/cdc_demo}
        \caption{Demographics}
    \end{subtable}
    \begin{subtable}{0.5\textwidth}
        \centering
        \input{tables/cdc_labs}
        \caption{Laboratory Results}
    \end{subtable}
       \begin{subtable}{0.5\textwidth}
        \centering
        \input{tables/cdc_exams}
        \caption{Physical Results}
    \end{subtable}
    \caption{Missing value distribution within each table/attribute in CDC NHANES 2013-2014 data}
    \label{table:nhanes-description} 
\end{table}

@@

For our purposes, we use a version of the data that has been 
pre-processed, but where many missing values remain. It is
worth noting that given the table design (where many attributes
are related), many of these missing values are to be
expected. The dataset we started with had 15,620 rows and 113
columns. We chose 17 attributes, which are outlined in~\Cref{table:fcc-description}, along with the percentage of missing values.

\begin{table}
   \centering
    \input{tables/fcc}
    \caption{Missing values in fCC Survey Data}
   \label{table:fcc-description} 
\end{table}

@@

\subsection{Queries}
We collect a set of queries (\Cref{fig:queries}) that we think are both representative and interesting
to plan. They could reasonably be written by a user in the course of data analysis.

@@

Queries 5-8 are on the fCC data. In~\Cref{q5}, we calculate the average income for
survey participants, based on their bootcamp attendance. In~\Cref{q6} estimates the
average age of women from the United States who participated. In~\Cref{q7}, we
calculate the average amount of money survey participants with student debt destined to
learning based on their school degree. In~\Cref{q8}, we join the fCC data with a
reference table provided by the World Bank which summarizes GDP per-capita across various
countries\cite{worldbank-data}. The query calculates the average GDP per-capita for
participants based on whether or not they attended a bootcamp.

@@

\begin{table*}
\centering
 \begin{subtable}{\linewidth}
  \input{queries-cdc}
  \caption{Queries on CDC data}
  \label{fig:queries-cdc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{queries-fcc}
 \caption{Queries on fCC data}
 \label{fig:queries-fcc}
 \end{subtable}
  \caption{Queries used in our experiments.}
  \label{fig:queries}
\end{table*}

@@

\Cref{fig:runtimes} shows a summary of the performance results. The first plot
shows average running times for \ProjectName{} plans in two configurations:
quality-optimized and runtime-optimized. The second plot shows, for comparison, the average
running times required for each query when performing imputation at the base tables.  In all
cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
with performance improvement on a factor of ten\todo{give range of performance}. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

\begin{figure}
\todo{change this to a single graph of X-times faster than base approach}
\begin{subfigure}{\linewidth}
\includegraphics[width=\columnwidth]{figures/running_times_imputedb.png}
\caption{Query runtimes with \ProjectName{}}
\label{subfig:project-runtime-queries}
\end{subfigure}
~
\begin{subfigure}{\linewidth}
\includegraphics[width=\columnwidth]{figures/running_times_base_tables.png}
\caption{Query runtimes with imputation on base tables}
\label{subfig:base-runtime-queries}
\end{subfigure}
\caption{Comparing \ProjectName{} runtimes for each query
    (\Cref{subfig:project-runtime-queries}) with the runtime for a query when base-table
    imputations are used (\Cref{subfig:base-runtime-queries})}
\label{fig:runtimes}
\end{figure}

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
\ProjectName{}'s query results when compared to running imputation on the base tables and
executing the query on the cleaned data. Each query ran for 200 iterations in both settings
and then results from the two approaches were paired up and compared tuple-wise. We average
tuple-wise absolute percentage deviations within each iteration of a query, and we report
this value averaged over all iterations.  We can see that optimizing for quality indeed
reduces the SMAPE of query results. In general, the SMAPE relative to the base-imputation
approach are low in all cases\todo{give actual value}, indicating that on-the-fly imputation produces similar
results to imputation at the base tables.

### 2017-02-27 paper/introduction.tex

% Context
Handling incorrect or dirty data is a complex and challenging problem for analysts\todo{add significant citations here}.
% Problem
One way in which a dataset can be dirty is for parts of it to be missing altogether.
Missing data, if handled naively, can cause analyses to be incorrect.
According to some estimates, data scientists spend between 50 and 80 percent of their time performing tasks such as collecting and cleaning datasets~\cite{data-science-cleaning}.

% Solution
To handle this problem, users may manually clean their dataset by performing some statistical analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
We have developed a system called \ProjectName{}, which is the first database to use a query optimization algorithm which allows imputation to occur on the fly, during query execution. \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}

>>>>

% Context
Handling incorrect or dirty data is a complex and challenging problem for analysts\todo{add significant citations here}.
% Problem
One way in which a dataset can be dirty is for parts of it to be missing altogether~\cite{kim2003}.
Missing data, if handled naively, can cause analyses to be incorrect~\cite{rubin1976}.
According to some estimates, data scientists spend between 50 and 80 percent of their time performing tasks such as collecting and cleaning datasets~\cite{data-science-cleaning}.

% Solution
To handle this problem, users may manually clean their dataset by performing some statistical analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
We have developed a system called \ProjectName{}, which is the first database to use a query optimization algorithm which allows imputation to occur on-the-fly, during query execution. \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}
